---
title: "Predict OVCA Subtypes"
author: aedin culhane
date: '`r Sys.Date()`'
output:
  html_document:
     toc: yes
bibliography: suppl.bib
---
```{r setup}
datadir = "../data"
srcdir= "../R"
source(file.path(srcdir, "ovca.R"))

library(made4)
library(curatedOvarianData)
library(caret)
```



## The TCGA molecular subtypes

The TCGA study [@TCGA2012] have described four high grade serous ovarian cancer molecular subtypes which are correlated to following the AOCS subtypes;

```{}
-------     -----------
  AOCS      TCGA
------      ------------
  C1        mesenchymal
  C2        immunoreative
  C4        differentiated *
  C5        proliferative
------      --------------
 
 * Note, C4 also correlated to immunoreative.
```


We downloaded a 100 gene signature of the TCGA molecular subtypes which was published as Supplementary Table 7 in the research article by Verhaak et al [@Verhaak2013]

Of these 100 genes, 90 were in the Bioconductor curatedOvarianData TCGA_eset [@Ganzfried2013]


```{r TCGA100Genes, warning=FALSE, message=FALSE}
require(curatedOvarianData)
data("TCGA_eset")

# Read TCGA 100 Gene Signature
tcgaGenes<-read.table(file.path(datadir,"23257362","Verhaak_13_CLOVAR100_Subtype.csv"), header=TRUE,  sep=",", as.is=TRUE)

boxplot(cbind(TCGA_eset$percent_normal_cells, TCGA_eset$percent_tumor_cells,TCGA_eset$percent_stromal_cells), ylim=c(0,100), names=c("N", "T", "TS"), col=c("grey", "yellow2", "green3"))

## Read the predicted subtypes of datasets
tcgaAnnotSubtypes<-read.table(file.path(datadir,"23257362", "Verhaak_13_suppl_SupplTab1.csv"), header=TRUE,  sep=",", skip=1, as.is=TRUE)
#table(tcgaAnnotSubtypes$DATASET)



# Filter the TCGA genes to those present in the curatedOVarianData TCGA_eset
print("Number of TCGA genes in the curatedOVarianData eset")
print(table(tcgaGenes[,1]%in%featureNames(TCGA_eset)))
tcgaGenes<-tcgaGenes[tcgaGenes[,1]%in%featureNames(TCGA_eset),]

```

The TCGA_eSet in the curated ovarian data contained 578 samples. Among these,  488 were in the discovery set, and we knew the tumor TCGA subtype classification.
 

```{r Compare_TCGA_Annot}
## Short code to compare the annotation of the 489 discovery subset from the first and second TCGA papers
## TCGA_489_UE.k4.txt was emailed by Roel Verhaak with the subtypes from the first TCGA paper

# Read TCGA tumor subtypes from the first TCGA paper.
tcgaSubtypesPaper1<-read.table(file.path(datadir,"23257362", "TCGA_489_UE.k4.txt"), sep="\t", header=TRUE, as.is=TRUE)

## Create a vector of the shortened Sample ID
tcgaSubtypesPaper1[,3]<-sub("-01[A,B,C,D]-0[1,2,3]R$", "", gsub("\\.", "\\-", tcgaSubtypesPaper1[,1]))
rownames(tcgaSubtypesPaper1) = tcgaSubtypesPaper1[,3]

tcgaAnnotSubtypes<-read.table(file.path(datadir,"23257362", "Verhaak_13_suppl_SupplTab1.csv"), header=TRUE,  sep=",", skip=1, as.is=TRUE)
#table(tcgaAnnotSubtypes$DATASET)

tt2<-tcgaAnnotSubtypes[grep("TCGA-discover",tcgaAnnotSubtypes$DATASET),]
rownames(tt2) = tt2$ID
TCGA_eset$SUBTYPE= tt2[TCGA_eset$unique_patient_ID,]$SUBTYPE
TCGA_eset$DATASET = "Validation"
TCGA_eset$DATASET[TCGA_eset$unique_patient_ID %in% tt2$ID]  ="DISCOVERY"
  
## Rownames are the same
#comparelists(rownames(tt2), rownames(tt1))
tcgaSubtypesPaper1<-tcgaSubtypesPaper1[rownames(tt2),]

## All TRUE, good annotation has not changed
if (!identical(tcgaSubtypesPaper1$k4, tt2$SUBTYPE)) print("TCGA annotation issue")
```

```{r TCGA_subtypes_TCGAgenes}
#TCGA_eset[tcgaGenes[,1],]
table(TCGA_eset$DATASET,TCGA_eset$SUBTYPE)
```

A hierarchical cluster analysis of the 90 molecular subtype genes, partitioned the TCGA tumor samples into subtypes. The Differentiated and Immunoreative clustered apart from the Mesenchymal and Proliferative subtypes.  

```{r TCGAHeatmap}
heatplot(TCGA_eset[as.character(tcgaGenes[,1]),], classvec=TCGA_eset$SUBTYPE,classvecCol =c("gray80", "grey20", "navy", "cyan"))
legend("topright", legend=levels(factor(TCGA_eset$SUBTYPE)), fill=c("gray80", "grey20", "navy", "cyan"), cex=0.5)
```

One of the samples in the Verhaak annotation was missing in TCGA_eSet."TCGA-13-0760" is absent in the curatedOvarianData package.   There are also 2 other samples ("TCGA-61-2611", "TCGA-61-2610" ) listed in the validation set that are missing.

```{r missing}
table(tcgaAnnotSubtypes[tcgaAnnotSubtypes[,1]%in%TCGA_eset$unique_patient_ID, "DATASET"])
tt2<-tcgaAnnotSubtypes[grep("TCGA",tcgaAnnotSubtypes$DATASET),]
comparelists(tt2[,1],TCGA_eset$unique_patient_ID)$Set.Diff
#comparelists(TCGA_eset$unique_patient_ID, tt2[,1])$Set.Diff
```




## The TCGA molecular subtypes: building a classifier

We built a SSP classifier of TCGA molecular subtypes using the 100 gene signature which were published as Supplementary Table 7 in the research article by Verhaak et al  [@Verhaak2013]

There were 90 of the 100 genes in the curatedOvarianData TCGA_eset and this subset of 90 were used to train the SSP classifier.  This performed with an overall self-validation accuracy of 89%. 


For a quick a dirty will use same approach as AOCS, then will build model
```{r tcgaSCM}
tcga_scm<-build.tcga() 
predictTCGA<-predict.tcga(eSet=TCGA_eset,  tcga.model=tcga_scm, corThres=0.7)

print(confusionMatrix(factor(predictTCGA$Predict), factor(TCGA_eset$SUBTYPE, levels=levels(factor((predictTCGA$Predict))))))
```

## TCGA molecular subtypes in microdissected tumor and stroma 

Apply the TCGA molecular subtype SSP classifier to the AOCS dataset.  

```{r tcgaPredict}
require(curatedOvarianData)
data("GSE9891_eset") # AOCS dataset
pt<-predict.tcga(GSE9891_eset, tcga.model=tcga_scm, corThres=0.7)
print("AOCS subtypes in microdissected tumors")
table(pt$Predict)
```

```{r plotPredictT}
tt<-plot.predict(pt)
title(main="Correlation of MGH tumors to TCGA subtyes")
```

*Spearman correlation of TCGA tumors to the centroids of TCGA subtpes *

## Comparing AOCS and TCGA classification.
The subtypes calls for the AOCS patients was published by Tothill, read the data

```{r AOCSsu}
supplClinAOCS<-read.csv(file.path(datadir, "18698038", "AOCSclinicaldata.csv"))
rownames(supplClinAOCS) = toupper(sub("\\.cel$", "",supplClinAOCS$X))
identical(rownames(supplClinAOCS),sampleNames(GSE9891_eset))
supplClinAOCS<-supplClinAOCS[sampleNames(GSE9891_eset),]

```

Cross-tabulation of TCGA subtypes and AOCS subtypes.
```{r xtab}
table(pt$Predict,supplClinAOCS$k.means.Group)
```




# Predicting OVCA molecular subtype background
Gene lists that predict AOCS molecular and TCGA subtypes were downloaded as online publication supplements  [@Tothill2008,@Verhaak2013] curated and submitted to GeneSigDB [@Culhane2012]. 
 
In this code, we downloaded the genelists and use  a simple single sample predictor (SSP) approachfor classification  [@HaibeKains2012; @Perou2000]. Tumors are classified as a molecular subtype, if a tumor had highest Spearman correlation to the median gene expression profile of the subtype, compared to the correlation to the median gene expression of the other subtypes.

## The AOCS molecular subtypes: building a classifier

The AOCS data is available as GSE9891_eset in the curatedOvarianData package, where features are GeneSymbols. In the case below, we downloaded it from GEO as I wanted probeIDs are features. However the map strategdy can be used to generate a classifier from GSE9891_eset


```{r buildaocs}
#Load AOCS (probeset) data
load(file.path(datadir,'AOCS_eSet.RData'))
tothill<-read.csv(file.path(datadir,"18698038", "18698038-TableS2.txt"), sep="\t", header=TRUE, as.is=TRUE)
AOCS_sig<-tothill[tothill[,1]%in%featureNames(aocs_eSet),]
aocs.sig=unique(AOCS_sig[,1])
if (!file.exists(file.path(datadir,"AOCS_Classifier_TothillGenes.Rda"))) {
  aocs.scm<-build.aocs(aocs_eSet, aocs.sig)
  save(aocs.scm,aocs.sig,file=file.path(datadir,"AOCS_Classifier_TothillGenes.Rda"))
}
```

```{r loadAOCSModel}
load(file.path(datadir,"AOCS_Classifier_TothillGenes.Rda"))
```

The Single Sample Predictor (SSP) classifier was trained using the published gene lists of the AOCS C1, C2, C4 and C5 molecular subtypes. The SSP identifies the subtype of a single tumor using a nearest centroid classifier and has been widely applied in molecular subtype classification in breast cancer  [@HaibeKains2012; @Perou2000].   It  predicted the AOCS molecular subtypes with an overall **self-validation accuracy** of 95%.  Sensitivity and specificityfor each subtype were is given below.

```{r aocsPred}
aocsPredict<-predict.aocs(eSet=aocs_eSet, aocs.model=aocs.scm, corThres=0.7)
#table(factor(aocsPredict$Predict), factor(aocs_eSet$k.means.Group))
#table(factor(aocs_eSet$MS), factor(aocs_eSet$k.means.Group))
print(confusionMatrix(factor(aocsPredict$Predict), factor(aocs_eSet$MS, levels=levels(factor(aocsPredict$Predict)))))


n=nrow(aocsPredict[order(aocs_eSet$MS),])

plot.predict(aocsPredict[order(aocs_eSet$MS),], reorderByClass = FALSE, colLab=aocs_eSet$k.means.Group[order(aocs_eSet$MS)], xlab="Tumors") 
title(main="Correlation of AOCS tumors to C1,C2,C4,C5 centroids")
```

In each case tumors are assigned to the subtype centroid to which it has highest correlation. Where there was no "winner", tumors has low correlation to all subtypes (cor<0.7), it was  unclassfied. 

```{r aocsClassSucc, fig.width=8, fig.height=10}
aocsPredict[aocsPredict$Predict=="Unclassified",]
par(mfrow=c(2,2))
for (i in colnames(aocsPredict)[1:4]) {boxplot(aocsPredict[aocsPredict$Predict==i,1:4], main=paste0("Classification of ",i ), las=3,ylim=c(0.4,1)); abline(h=0.7,lty=2,col="red" )}
```

It can be seens that C1 and C2 tumors, showed weaker correlation but >0.7 correlation to C1,C2 and C4 tumors. C1 and C2 tumors had lowest correlation to C5 tumors. Consistent with this, C5 were easier to classify, and mostly had <0.7 correlation to C1,C2,C4. This was reflect in the confusious matrix above, were C5 tumors were classified with the higher Balanced Accuracy (0.98), Sensitivity (0.97) and Specificity (0.99).

*Spearman correlation of `r n` AOCS tumor samples to the centroid predictor for molecular subtype C1 (light grey), C2 (dark grey), C4 (navy), C5 (cyan).*

## AOCS molecular subtypes in curatedOvarianData


Example of applying classifiers to public data

```{r CuratedOvarianData, eval=FALSE}
require(curatedOvarianData)
testaocsPredict<-predict.aocs(Tumors, aocs.model=aocs.scm,corThres=0.7)

#Plot of results
plot.predict(testaocsPredict, colLab=Tumors$alt_sample_name, xlab="Tumors") 

```


